# Data

## Data Processing

### Weather Data Preprocessing

```{r}
#| label: weather-preprocessing

# Set file paths
weather_temp_path <- here("data", "monthly_nyc_weather.csv")
weather_rain_path <- here("data", "monthly_nyc_rain.csv")

# Function to preprocess weather data
preprocess_weather_data <- function() {
  # Load temperature and rainfall data
  temp_df <- read_csv(weather_temp_path)
  rain_df <- read_csv(weather_rain_path)
  
  # Convert Month format to date
  temp_df <- temp_df %>%
    mutate(Date = as.Date(paste0(Month, "01"), format = "%Y%m%d"))
  
  rain_df <- rain_df %>%
    mutate(Date = as.Date(paste0(Month, "01"), format = "%Y%m%d"))
  
  # Rename columns for clarity
  temp_df <- temp_df %>%
    rename(Temperature = Value, TempAnomaly = Anomaly)
  
  rain_df <- rain_df %>%
    rename(Rainfall = Value, RainAnomaly = Anomaly)
  
  # Merge temperature and rainfall data
  weather_df <- temp_df %>%
    inner_join(rain_df, by = c("Date"))
  
  # Extract date components for analysis
  weather_df <- weather_df %>%
    mutate(
      Year = year(Date),
      Month = month(Date),
      # Create season feature
      Season = case_when(
        Month %in% c(12, 1, 2) ~ "Winter",
        Month %in% c(3, 4, 5) ~ "Spring",
        Month %in% c(6, 7, 8) ~ "Summer",
        TRUE ~ "Fall"
      )
    )
  
  return(weather_df)
}

# Process weather data
weather_df <- preprocess_weather_data()

# Display the first few rows
head(weather_df)
```

### Traffic Data Preprocessing

```{r}
#| label: traffic-preprocessing
#| eval: false

traffic_path <- here("data", "Automated_Traffic_Volume_Counts_20250505.csv")

# Function to preprocess traffic data
preprocess_traffic_data <- function(sample_size = NULL) {
  # Load traffic data with sampling if needed due to size
  if (!is.null(sample_size)) {
    traffic_df <- read_csv(traffic_path, n_max = sample_size)
  } else {
    traffic_df <- read_csv(traffic_path)
  }
  
  # Handle missing values
  traffic_df <- traffic_df %>%
    filter(!is.na(Vol))
  
  # Create datetime column and extract components
  traffic_df <- traffic_df %>%
    mutate(
      DateTime = as.POSIXct(paste(Yr, M, D, HH, MM, sep = "-"), format = "%Y-%m-%d-%H-%M"),
      Year = year(DateTime),
      Month = month(DateTime),
      Day = day(DateTime),
      DayOfWeek = wday(DateTime) - 1,  # 0 = Sunday, 6 = Saturday
      Hour = hour(DateTime),
      
      # Create features for time of day
      TimeOfDay = case_when(
        Hour >= 6 & Hour < 10 ~ "Morning",
        Hour >= 10 & Hour < 16 ~ "Midday",
        Hour >= 16 & Hour < 20 ~ "Evening",
        TRUE ~ "Night"
      ),
      
      # Create weekday/weekend feature
      IsWeekend = if_else(DayOfWeek >= 5, 1, 0)
    )
  
  return(traffic_df)
}

# Process traffic data (with sampling due to large file size)
traffic_df <- preprocess_traffic_data(sample_size = 100000)

# Show summary
summary(traffic_df)
```

### Emergency Response Data Preprocessing

```{r}
#| label: emergency-preprocessing

emergency_path <- here("data", "911_Open_Data_Local_Law_119_20250505.csv")

# Function to preprocess emergency response data
preprocess_emergency_data <- function() {
  emergency_df <- read_csv(emergency_path)
  
  # Extract date from 'Month Name' column
  emergency_df <- emergency_df %>%
    mutate(
      Date = as.Date(paste0("01 ", `Month Name`), format = "%d %Y / %m"),
      Year = year(Date),
      Month = month(Date)
    )
  
  return(emergency_df)
}

# Process emergency data
emergency_df <- preprocess_emergency_data()

# Display the structure
glimpse(emergency_df)
```

## Exploratory Data Analysis


```{r}
#| label: weather-plots
#| fig.width: 10
#| fig.height: 6

# 1. Weather patterns over time
ggplot(weather_df, aes(x = Date, y = Temperature)) +
  geom_line() +
  labs(
    title = "Monthly Temperature in NYC (2013-2025)",
    x = "Date",
    y = "Temperature (F)"
  ) +
  theme_minimal()

ggplot(weather_df, aes(x = Date, y = Rainfall)) +
  geom_line() +
  labs(
    title = "Monthly Rainfall in NYC (2013-2025)",
    x = "Date",
    y = "Rainfall (inches)"
  ) +
  theme_minimal()
```

```{r}
#| label: traffic-plots
#| fig.width: 10
#| fig.height: 6
#| eval: false

# Traffic volume by time of day
traffic_df %>%
  group_by(TimeOfDay) %>%
  summarize(MeanVolume = mean(Vol, na.rm = TRUE)) %>%
  mutate(TimeOfDay = factor(TimeOfDay, levels = c("Morning", "Midday", "Evening", "Night"))) %>%
  ggplot(aes(x = TimeOfDay, y = MeanVolume)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Traffic Volume by Time of Day",
    x = "Time of Day",
    y = "Average Volume"
  ) +
  theme_minimal()

# Traffic volume by borough
traffic_df %>%
  group_by(Boro) %>%
  summarize(MeanVolume = mean(Vol, na.rm = TRUE)) %>%
  arrange(desc(MeanVolume)) %>%
  ggplot(aes(x = reorder(Boro, MeanVolume), y = MeanVolume)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Traffic Volume by Borough",
    x = "Borough",
    y = "Average Volume"
  ) +
  theme_minimal() +
  coord_flip()
```

```{r}
#| label: emergency-plots
#| fig.width: 10
#| fig.height: 6
#| eval: false

# Emergency response times by borough
ggplot(emergency_df, aes(x = Borough, y = `Response Times`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Emergency Response Times by Borough",
    x = "Borough",
    y = "Response Time (seconds)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Data Integration

```{r}
#| label: data-integration
#| eval: false

# Function to integrate datasets
integrate_datasets <- function(traffic_df, weather_df, emergency_df) {
  # Aggregate traffic data to daily level
  daily_traffic <- traffic_df %>%
    group_by(Year, Month, Day, Boro) %>%
    summarize(Vol = mean(Vol, na.rm = TRUE), .groups = "drop") %>%
    mutate(Date = as.Date(paste(Year, Month, Day, sep = "-")))
  
  # Expand weather data to daily
  weather_expanded <- weather_df %>%
    select(Date, Temperature, Rainfall, TempAnomaly, RainAnomaly, Season) %>%
    mutate(
      month_start = floor_date(Date, "month"),
      month_end = ceiling_date(Date, "month") - days(1)
    )
  
  # Create daily weather records
  daily_weather_records <- list()
  
  for (i in 1:nrow(weather_expanded)) {
    row <- weather_expanded[i, ]
    date_range <- seq(row$month_start, row$month_end, by = "day")
    
    for (date in date_range) {
      daily_record <- row
      daily_record$Date <- date
      daily_weather_records[[length(daily_weather_records) + 1]] <- daily_record
    }
  }
  
  daily_weather <- bind_rows(daily_weather_records) %>%
    select(-month_start, -month_end)
  
  # Merge daily traffic with daily weather
  merged_df <- daily_traffic %>%
    inner_join(daily_weather, by = "Date") %>%
    mutate(
      DayOfWeek = wday(Date) - 1,
      IsWeekend = if_else(DayOfWeek >= 5, 1, 0)
    )
  
  return(merged_df)
}

# Integrate datasets
integrated_data <- integrate_datasets(traffic_df, weather_df, emergency_df)

# Save processed data
write_csv(weather_df, here("data", "processed_weather_data.csv"))
write_csv(traffic_df, here("data", "processed_traffic_data.csv"))
write_csv(emergency_df, here("data", "processed_emergency_data.csv"))
write_csv(integrated_data, here("data", "integrated_data.csv"))

# Display the structure of the integrated dataset
glimpse(integrated_data)
```

## Feature Engineering

```{r}
#| label: feature-engineering
#| eval: false

# Function to engineer features
engineer_features <- function(df) {
  # One-hot encode categorical variables
  df <- df %>%
    mutate(
      Season_Spring = if_else(Season == "Spring", 1, 0),
      Season_Summer = if_else(Season == "Summer", 1, 0),
      Season_Fall = if_else(Season == "Fall", 1, 0),
      Season_Winter = if_else(Season == "Winter", 1, 0)
    )
  
  # Create borough dummy variables
  boroughs <- unique(df$Boro)
  for (borough in boroughs) {
    col_name <- paste0("Boro_", borough)
    df[[col_name]] <- if_else(df$Boro == borough, 1, 0)
  }
  
  # Add lagged features (previous day's traffic volume)
  for (borough in boroughs) {
    col_name <- paste0("Boro_", borough)
    lag_col_name <- paste0(borough, "_prev_day_vol")
    
    # Filter for this borough and sort by date
    boro_data <- df %>%
      filter(!!sym(col_name) == 1) %>%
      arrange(Date)
    
    # Create lagged volume
    boro_data <- boro_data %>%
      mutate(!!lag_col_name := lag(Vol))
    
    # Update the original dataframe
    df <- df %>%
      left_join(
        boro_data %>% select(Date, Boro, !!lag_col_name),
        by = c("Date", "Boro")
      )
  }
  
  # Fill NA values from lagged features with mean values
  for (col in names(df)) {
    if (any(is.na(df[[col]]))) {
      col_mean <- mean(df[[col]], na.rm = TRUE)
      df[[col]] <- if_else(is.na(df[[col]]), col_mean, df[[col]])
    }
  }
  
  return(df)
}

# Engineer features
engineered_data <- engineer_features(integrated_data)

# Save engineered data
write_csv(engineered_data, here("data", "engineered_data.csv"))

# Summary of engineered features
summary(engineered_data)
```

## Summary

1. **Loaded and cleaned** three distinct datasets: traffic volumes, weather patterns, and emergency response times
2. **Created time-based features** including time of day, day of week, and seasonal indicators
3. **Performed exploratory visualizations** to understand traffic patterns by time and location
4. **Integrated the datasets** to create a comprehensive view of NYC traffic patterns
5. **Engineered additional features** to improve the model's predictive capabilities

The prepared data will be called in the next chapter to develop the models and analysis.
